---
title: "Types of storms and other severe weather events that have greatest consequences on human health and economy in the USA"
author: "Pierre Nordmann"
date: "23 May 2017"
output: html_document
---

## Synopsis
The purpose of this document is to present the types of storms and other severe weather events that have the greatest consequences on human health and economy in the USA.
The data used for this analysis is published by the National Climatic Data Center, that itself receives the data from the National Weather Service. The events in the database start in the year 1950 and end in November 2011. 

There is some data preparation in this analysis: labels of event type are modified, in order to have group together event types in a relevant manner.

3 indicators have been caculated:

* number of persons injured per event type
* number of fatalities per event type
* sum of amounts of crop damage and property damage

The results are that:

* Tornadoes, heat and flash flood cause the most fatalities
* Tornadoes are the event that cause the most injuries by far
* Flood, Hurricane/typhoon and tornadoes cause the most economic damages


## Data processing
Let's load libraries.
```{r libraries, results="hide", echo=TRUE}
library(ggplot2)
library(scales)
```


Let's download the data.
The data used for this analysis is published by the National Climatic Data Center, that itself receives the data from the National Weather Service. The events in the database start in the year 1950 and end in November 2011. 
<br/>Let's create a data frame, so that it can be processed.
```{r data_loading, cache=TRUE, results="hide", echo=TRUE}
download.file(url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
              destfile = "./repdata%2Fdata%2FStormData.csv.bz2")

Storm_Data <- read.csv(file = "./repdata%2Fdata%2FStormData.csv.bz2")
```
<br/>  

The Storm Data event table mentionned at page 6 of the [Storm Data dataset notice](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) is not 100% consistent with the EVTYPE column of the dataset.
The operations below correct this issue.

```{r data_preparation, results="hide", echo=TRUE}
# Let's rename Event type for the labels of Event type that look very similar
Storm_Data$EVTYPE[Storm_Data$EVTYPE == "TSTM WIND"] <- "THUNDERSTORM WIND"
Storm_Data$EVTYPE[Storm_Data$EVTYPE == "THUNDERSTORM WINDS"] <- "THUNDERSTORM WIND"
Storm_Data$EVTYPE[Storm_Data$EVTYPE == "URBAN FLOOD"] <- "FLOOD"
Storm_Data$EVTYPE[Storm_Data$EVTYPE == "FLASH FLOODING"] <- "FLASH FLOOD"
Storm_Data$EVTYPE[Storm_Data$EVTYPE == "RIVER FLOOD"] <- "FLOOD"
Storm_Data$EVTYPE[Storm_Data$EVTYPE == "MARINE TSTM WIND"] <- "MARINE THUNDERSTORM WIND"
Storm_Data$EVTYPE[Storm_Data$EVTYPE == "WILD/FOREST FIRE"] <- "WILDFIRE"
```
<br/>  

The code below sums fatalities per event type, orders the event types per descending number of fatalities, and keeps only the top 20 event types.
```{r Fatalities, results="hide", echo=TRUE}
# Let's sum fatalities and injuries for each event type
Storm_Data_Fatalities_per_eventtype<- aggregate(Storm_Data$FATALITIES, by = Storm_Data[c('EVTYPE')], sum)

Fatalities <- head(Storm_Data_Fatalities_per_eventtype[order(-Storm_Data_Fatalities_per_eventtype$x),],20)
```
<br/>

The code below sums injuries per event type, orders the event types per descending number of injuries, and keeps only the top 20 event types.
```{r Injuries, results="hide", echo=TRUE}
Storm_Data_Injuries_per_eventtype<- aggregate(Storm_Data$INJURIES, by = Storm_Data[c('EVTYPE')], sum)

Injuries <- head(Storm_Data_Injuries_per_eventtype[order(-Storm_Data_Injuries_per_eventtype$x),],20)
```
<br/>  
Regarding the economic damages, the following operations have to be performed.

1. Unit and values are in two different columns. A new column with the value, including the unit, should be created.
2. Crop damages and property damages are in different columns. They should be added into a new column.
3. Sum damages per event type, order the event types per descending number of damages, and keeps only the top 20 event types.

```{r Damages, results="hide", echo=TRUE}
# In order to compute the data about damages,we have to create a column that would multiply PROPDMG by the value of PROPDMGEXP
# if h or H, multiply by 100, if k or K, multiply by 1000, if m or M, by 1000000, if B, by 10 exp9, else do not multiply

Storm_Data$PROPDMGEXP2[grepl("[hH]",Storm_Data$PROPDMGEXP)] <- 100
Storm_Data$PROPDMGEXP2[grepl("[kK]",Storm_Data$PROPDMGEXP)] <- 1000
Storm_Data$PROPDMGEXP2[grepl("[mM]",Storm_Data$PROPDMGEXP)] <- 1000000
Storm_Data$PROPDMGEXP2[grepl("[bB]",Storm_Data$PROPDMGEXP)] <- 1000000000
Storm_Data$PROPDMGEXP2[grepl("[mM]",Storm_Data$PROPDMGEXP) + grepl("[bB]",Storm_Data$PROPDMGEXP) + grepl("[kK]",Storm_Data$PROPDMGEXP)+ grepl("[hH]",Storm_Data$PROPDMGEXP) == 0 ] <- 1

Storm_Data$CROPDMGEXP2[grepl("[hH]",Storm_Data$CROPDMGEXP)] <- 100
Storm_Data$CROPDMGEXP2[grepl("[kK]",Storm_Data$CROPDMGEXP)] <- 1000
Storm_Data$CROPDMGEXP2[grepl("[mM]",Storm_Data$CROPDMGEXP)] <- 1000000
Storm_Data$CROPDMGEXP2[grepl("[bB]",Storm_Data$CROPDMGEXP)] <- 1000000000
Storm_Data$CROPDMGEXP2[grepl("[mM]",Storm_Data$CROPDMGEXP) + grepl("[bB]",Storm_Data$CROPDMGEXP) + grepl("[kK]",Storm_Data$CROPDMGEXP) + grepl("[hH]",Storm_Data$CROPDMGEXP)== 0 ] <- 1

Storm_Data$PROPDMG2 <- Storm_Data$PROPDMG * Storm_Data$PROPDMGEXP2
Storm_Data$CROPDMG2 <- Storm_Data$CROPDMG * Storm_Data$CROPDMGEXP2
Storm_Data$Damage <- Storm_Data$PROPDMG2 + Storm_Data$CROPDMG2

Damage_per_eventtype <- aggregate(Storm_Data$Damage, by = Storm_Data[c('EVTYPE')], sum)
Damage_per_eventtype$x <- Damage_per_eventtype$x / 1000000000
Damages <- head(Damage_per_eventtype[order(-Damage_per_eventtype$x),],20)
```
<br/>  

## Results
Tornadoes, heat and flash flood cause the most fatalities.

```{r results_fatalities, fig.width=10, fig.height=6, include = TRUE, echo = TRUE}
p <-  ggplot(Fatalities, aes(x = reorder(EVTYPE,-x),y = x))
p <- p+ geom_bar(stat  = "identity")
p <- p+ ggtitle("Top 20 Event types in terms of Fatalities")
p <- p+ geom_text(aes(label=comma(x), vjust=-0.5, size=2.5))
p <- p+scale_y_continuous(limits=c(0,6000), labels = comma)
p <- p+ ylab("Number of fatalities") + xlab("Event type")
p <- p+ theme(legend.position="none", plot.title = element_text(size = 15),axis.title = element_text(size = 13),axis.text.y = element_text(size=13), axis.text.x = element_text(angle=90,size=13))
p
```
<br/><br/><br/>

Tornadoes are the event that cause the most injuries by far.
```{r results_injuries, fig.width=10, fig.height=6, include = TRUE, echo = TRUE}
p <-  ggplot(Injuries, aes(x = reorder(EVTYPE,-x),y = x))
p <- p+ geom_bar(stat = "identity")
p <- p+ ggtitle("Top 20 Event types in terms of Injuries")
p <- p+ geom_text(aes(label=comma(x), vjust=-0.5, size=2.5))
p <- p+scale_y_continuous(limits=c(0,100000), labels = comma )
p <- p+ ylab("Number of injuries") + xlab("Event type")
p <- p+ theme(legend.position="none", plot.title = element_text(size = 15),axis.title = element_text(size = 13),axis.text.y = element_text(size=13), axis.text.x = element_text(angle=90,size=13))
p
```
<br/><br/><br/>
Flood, Hurricane/typhoon and tornadoes cause the most economic damages (crop and property).
```{r results_damages, fig.width=10, fig.height=6, include = TRUE, echo = TRUE}
p <-  ggplot(Damages, aes(x = reorder(EVTYPE,-x),y = x))
p <- p+ geom_bar(stat = "identity")
p <- p+ ggtitle("Top 20 Event types in terms of Crop and Property Damages")
p <- p+ geom_text(aes(label=sprintf("%1.0f", round(x,digits = 0)), vjust=-0.5, size=2.5))
p <- p+scale_y_continuous(limits=c(0,200), labels = comma )
p <- p+ ylab("Billion Dollars") + xlab("Event type")
p <- p+ theme(legend.position="none", plot.title = element_text(size = 15),axis.title = element_text(size = 13),axis.text.y = element_text(size=13), axis.text.x = element_text(angle=90,size=13))
p
```



